### 2022-07-29 manuscript.Rmd
revision

authornote        : "Word count: 7956 words"
note              : "This working paper has not yet been peer-reviewed."
keywords          : "cooperation, conformity, punishment, behavioural economics, political ideology"
wordcount         : "7956 words"
bibliography      : ["r-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
appendix          : "appendix.Rmd"
---

@@

Humans differ profoundly in their views on political issues like income redistribution, taxation, welfare, abortion, and religious instruction in schools. With an increasing focus on political polarisation and the consequent challenges of finding common ground on a host of pressing issues [@Neal2018; @PewResearchCenter2014], understanding the sources of this variation is more pressing than ever. One prominent explanation for variation in political views is that liberals and conservatives differ fundamentally in psychological traits and social preferences [@Hibbing2014; @Jost2003]. For example, research suggests that, compared to liberals, conservatives exhibit increased sensitivity to threats [@Hibbing2014], personal needs for closure [@Jost2003], and intolerance of uncertainty [@Jost2017], producing variation in support for policies that promote order, structure, and stability. By contrast, other evidence suggests that liberals and conservatives do not differ psychologically [@Brandt2014; @Crawford2017a; @Crawford2017b] and that variation in political views between liberals and conservatives instead reflects expressions of self-interest based on socio-demographic factors, identity, and political coalitions [@Weeden2014].

The debate about the existence of liberal-conservative psychological differences has largely treated political ideology as a unidimensional spectrum from left to right [but see @Crawford2017b]. But decades of interdisciplinary work indicate that political ideology is best described along two dimensions [@Claessens2020; @Duckitt2009; @Feldman2014; @Saucier2000; @Schwartz1992]. The first dimension captures views on issues like taxation, welfare, capitalism, and big business [@Feldman2014], and has been variously labelled as economic conservatism, Social Dominance Orientation (SDO), individualising moral foundations, equality values, and egalitarianism [@Claessens2020; @Duckitt2009; @Graham2009; @Rokeach1973]. The second dimension captures views on issues like abortion, patriotism, sexual morality, and criminal justice [@Feldman2014], and has been variously labelled as social conservatism, Right Wing Authoritarianism (RWA), binding moral foundations, freedom values, and group loyalty [@Claessens2020; @Duckitt2009; @Graham2009; @Rokeach1973]. These two dimensions have repeatedly emerged from independent lines of research in political psychology [@Duckitt2009], moral psychology [@Graham2009], and cross-cultural psychology [@Ashton2005]. While the two dimensions of political ideology tend to positively covary in Western societies, suggesting that they are not orthogonal [@Duckitt2009], they are distinct and can vary independently, being negatively correlated with one another in many societies around the globe [@Malka2019].

Building on previous evolutionary approaches to politics and human sociality [@Haidt2012; @Hibbing2013; @Petersen2015], we recently proposed that the dual dimensions of political ideology are a product of two key trade-offs inherent to the evolution of human group living [@Claessens2020]: a trade-off between cooperative egalitarianism and competitive hierarchy, and a trade-off between individualism and group conformity. In line with approaches claiming that psychological and behavioural differences exist across the political landscape, one prediction of the dual evolutionary framework is that variation in general behavioural preferences for cooperation and group conformity will differentially predict the two dimensions of ideology in modern humans [@Claessens2020]^[We define cooperative behaviour as behaviour that provides a benefit to another individual at a personal cost [@West2007] and conformist behaviour as reliance on socially acquired information and adherence to and enforcement of social norms within cultural groups.]. Under this framework, an increased general behavioural preference for cooperation beyond close kin results in greater support for policies that promote large-scale impersonal cooperation and a more egalitarian distribution of resources, such as income redistribution, taxation, welfare, and pro-environmentalism. An increased general behavioural preference for group conformity results in greater support for policies that promote in-group conformity, norm enforcement, and the interests of the in-group, such as religious instruction in schools, capital punishment, and military spending.

If general behavioural preferences for cooperation and group conformity do underlie differences in political opinion in modern humans, such differences should be manifest in preferences measured by incentivised behavioural tasks. Incentivised behavioural tasks (i.e. decision-making tasks involving real money) allow researchers to study variation in enduring social preferences without succumbing to issues of social desirability and experimenter demand that plague self-report methods [@Chaudhuri2009; @Gawronski2015; @Pisor2020]. Decades of work using incentivised behavioural tasks have revealed individual differences in both cooperative behavioural preferences [@Kuhlman1975; @Kurzban2001; @Chaudhuri2011] and conformist behavioural preferences [@Molleman2019; @Toelch2014; @Efferson2008; @Molleman2014] in humans.

@@

Here, we extend prior work by combining a large suite of incentivised behavioural tasks with comprehensive survey data to test whether and to what extent cooperative and conformist behavioural preferences covary with measures of the two dimensions of political ideology. Across two pre-registered studies, we asked participants to complete a series of incentivised behavioural tasks measuring cooperative preferences, such as trust, altruism, and collective action, and conformist preferences, such as norm-enforcing punishment, social information use, and rule following. In Study 1, we used a battery of one-shot anonymous economic games, selected because they have previously been shown to capture two dimensions of behaviour, one reflecting a "cooperative phenotype" and another labelled "norm-enforcing punishment" [@Peysakhovich2014]. In Study 2, we extended this suite of games to include incentivised tasks capturing two further elements of conformist behaviour: social information use and rule following.

@@

Participants completed eight economic games, conducted online in real-time using oTree software [code available at https://osf.io/dwx8g/; @Claessens2021; @Chen2016]. In an attempt to replicate previous work, these games are largely identical to those from @Peysakhovich2014.

@@

All games involved one-shot decisions between multiple anonymous players. Such one-shot anonymous games remove the well-documented effects of communication and known players on behaviour [@Chaudhuri2009; @Chaudhuri2011] and avoid potential confounds arising from repeated play, given that repeated game play may result in "super-games" and consequent cooperation even in finite repetitions [@Kreps1982]. The strategy method was used to measure responses in all possible roles, an approach which avoids feedback from other players and elicits similar behaviour to sequential play [@Brandts2000].

@@

Having confirmed the existence of a two-factor structure underlying gameplay, we next fitted pre-registered structural equation models to test the relationships between the two latent variables and political ideology. We simultaneously regressed the cooperation and punishment latent variables onto mean scores of six Likert-scale items for SDO (Cronbach's $\alpha$ = `r printnum(readd(alphaSDO))`) and RWA ($\alpha$ = `r printnum(readd(alphaRWA))`). In line with our second pre-registered hypothesis, we found that SDO significantly negatively predicted the cooperation latent variable with a small-to-medium effect size (standardised $\beta$ = `r printnum(standardizedSolution(sem1)[19,4])`, unstandardised _b_ = `r printnum(parameterEstimates(sem1)[19,5])`, 95% confidence interval [`r printnum(parameterEstimates(sem1)[19,9])`, `r printnum(parameterEstimates(sem1)[19,10])`], _p_ `r printp(parameterEstimates(sem1)[19,8])`, semi-partial _r_ = `r printnum(r_SDO_coop)`; Figure \@ref(fig:coopPunPlot)a). SDO also significantly positively predicted the punishment latent variable with a small effect size, which we did not hypothesise ($\beta$ = `r printnum(standardizedSolution(sem1)[21,4])`, _b_ = `r printnum(parameterEstimates(sem1)[21,5], digits=3)` [`r printnum(parameterEstimates(sem1)[21,9], digits=3)`, `r printnum(parameterEstimates(sem1)[21,10], digits=3)`], _p_ = `r printp(parameterEstimates(sem1)[21,8])`, semi-partial _r_ = `r printnum(r_SDO_pun)`), though the magnitude of this association was not as strong as with the cooperation latent variable (difference in absolute unstandardized coefficients = `r printnum(parameterEstimates(sem1)[59,5], digits=2)`, SE = `r printnum(parameterEstimates(sem1)[59,6], digits=3)`, p `r printp(parameterEstimates(sem1)[59,8])`). In contrast with our third pre-registered hypothesis, we found no significant relationship between RWA and the punishment latent variable ($\beta$ = `r printnum(standardizedSolution(sem1)[22,4])`, _b_ = `r printnum(parameterEstimates(sem1)[22,5], digits=3)` [`r printnum(parameterEstimates(sem1)[22,9], digits=3)`, `r printnum(parameterEstimates(sem1)[22,10], digits=3)`], _p_ = `r printp(parameterEstimates(sem1)[22,8])`, semi-partial _r_ = `r printnum(r_RWA_pun)`; Figure \@ref(fig:coopPunPlot)b). RWA was also unrelated to the cooperation latent variable ($\beta$ = `r printnum(standardizedSolution(sem1)[20,4])`, _b_ = `r printnum(parameterEstimates(sem1)[20,5], digits=2)` [`r printnum(parameterEstimates(sem1)[20,9], digits=2)`, `r printnum(parameterEstimates(sem1)[20,10], digits=2)`], _p_ = `r printp(parameterEstimates(sem1)[20,8])`, semi-partial _r_ = `r printnum(r_RWA_coop)`). This pattern of results held when additionally controlling for age, gender, ethnicity, education, socio-economic status, local deprivation, and religiosity (Supplementary Table \@ref(tab:semTable)).

@@

To further investigate the relationships between political ideology and gameplay, we conducted a series of exploratory linear regressions with mean SDO and RWA scores as outcome variables and individual game decisions as predictors, accounting for multiple comparisons and statistically controlling for demographics and game comprehension. We found that most cooperative decisions were negatively associated with SDO (Supplementary Figure \@ref(fig:indGamesPlotSDO)). Holding all other variables constant, individuals with higher SDO scores gave less in the Dictator Game, returned less in the Trust Game, contributed less in the Public Goods Game, did not coordinate in the Stag Hunt Game, offered less in the Ultimatum Game, and were more likely to take in the Third-Party Punishment Game. The effect sizes for these relationships were small, but consistent across games (semi-partial _r_ = `r min(rs)`-`r max(rs)`). Contrary to our predictions, SDO was also positively related to both "social" and "antisocial" punishment decisions in the Second-Party Punishment Game and Stag Hunt Game with Punishment (i.e. costly punishment of defectors and cooperators, respectively). Fewer individual game decisions predicted RWA (Supplementary Figure \@ref(fig:indGamesPlotRWA)). Holding all other variables constant, individuals with higher RWA scores were more likely to take in the Third-Party Punishment Game and engaged in greater anti-social punishment in the Second-Party Punishment Game and the Stag Hunt Game with Punishment.

@@

We did not find the predicted association between the punishment latent variable and RWA, nor did we find any link between the punishment latent variable and socially conservative policy views. One explanation for this lack of relationship is that, while authoritarians might desire the sanctioning of norm violators, they would rather this punishment come from a third-party institution that they perceive to be legitimate than from an interpersonal encounter. Indeed, RWA items refer to sanctioning from strong leaders or governments, but rarely from individuals themselves [@Duckitt2009]. Another explanation is that the punishment games used in this study do not reflect general group conformist preferences as expected. As has been suggested previously [@Price2002; @Raihani2019], punishment in economic games may more closely track a competitive motive to increase one's relative payoffs over others, especially when the economic games are one-shot games with anonymous others in which there is no possibility for signalling, reputation building, or future behaviour modification.

@@

The materials for Study 2 were mostly identical to Study 1. As in Study 1, eight incentivised behavioural tasks were conducted online in real-time using oTree software [code available at https://osf.io/dwx8g/; @Claessens2021; @Chen2016]. The strategy method was used to elicit responses in all possible roles, and participants played for points which were converted to New Zealand dollars at the same rate as in Study 1 (1 point = $0.035). The six tasks from previous work [@Peysakhovich2014] --- the Dictator Game, Trust Game, Public Goods Game, Ultimatum Game, Third-Party Punishment Game, and Second-Party Punishment Game --- were measured again (see Supplementary Table \@ref(tab:testRetestTable) for test-retest reliability). However, in Study 2, the coordination games were replaced with two different tasks.

The first task was the Berlin Estimate AdjuStment Task [BEAST; @Molleman2019], a task designed to measure individual differences in conformity in the form of social information use. In this task, over five rounds, participants are required to report the number of animals present in briefly displayed images (see Supplementary Figure \@ref(fig:beastImages)). There are two parts to each round. In the first part, the image is displayed for six seconds and participants give their initial estimate of the number of animals ($E_1$). In the second part, participants are then made aware of a previous participant's estimate for the same image ($s$) and then give a second estimate ($E_2$). Previous participants' estimates are determined by searching a database of real estimates from 100 previous participants sampled by @Molleman2019. The computer searches for an estimate in this database that is (1) in the direction of the true answer (e.g. if the participant guesses 50 and the true answer is 60, social information will be higher than 50) and (2) deviates from the participant's first estimate according to a parameter $\Delta$: $s = E_1 * (1 \pm \Delta)$. Following the original paper, this $\Delta$ parameter varied across the five rounds, taking values of 0.25, 0.15, 0.20, 0.15, and 0.25 respectively. Participants were paid for their accuracy by randomly choosing a round and determining the distance between their estimate and the correct answer: 100 points - (distance * 5).

@@

In sum, we have shown clear differences in social preferences across the two dimensions of political ideology. We found that a general cooperative preference covaried with SDO and economic policy views, and informational conformity covaried with RWA and social policy views. These psychological differences were largely consistent with predictions of the dual evolutionary framework of political ideology [@Claessens2020]. The predictions not supported by the data suggest that the relationship between informational conformity and RWA and social policy views may not extend to conformity in the form of norm-enforcement and rule following, although demonstrating this conclusively will require further research using alternative behavioural measures. More generally, our results challenge claims that there are no fundamental psychological differences between liberals and conservatives, adding to the catalogue of traits that have been found to differ across the political landscape. Moreover, by characterising the very different social preferences underlying both ideological dimensions, our findings highlight the importance of work that moves beyond unidimensional scales to elucidate the dual evolutionary foundations of political ideology.

@@

This work was supported by a Royal Society of New Zealand Marsden grant (UOA1711).

@@

# (APPENDIX) Appendix {-}

```{r child = "appendix.Rmd"}
```

>>>>

authornote        : "Word count: 8323 words"
#note              : "This working paper has not yet been peer-reviewed."
keywords          : "cooperation, conformity, punishment, behavioural economics, political ideology"
wordcount         : "8323 words"
bibliography      : ["r-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
appendix          : "appendix.Rmd"
---

@@

Humans differ profoundly in their views on political issues like income redistribution, taxation, welfare, abortion, and religious instruction in schools. With an increasing focus on political polarisation and the consequent challenges of finding common ground on a host of issues [@Neal2018; @PewResearchCenter2014], understanding the sources of this variation is more pressing than ever.

The extent to which individuals’ political views reflect a predictable ideological structure, and the psychological explanations for ideological differences, are longstanding questions in social psychology that have received renewed attention in recent years [@Jost2006]. Political ideology is defined as a set of stable, interrelated beliefs and attitudes that organise views on political and social issues [@Jost2009]. Traditionally, political ideology has been conceptualised as a unidimensional spectrum that varies from liberalism on the left to conservatism on the right [@Claessens2020; @Adorno1950]. Under this standard view, liberals favour equality and social change, whereas conservatives value hierarchy and traditionalism.

However, interdisciplinary research over the past five decades has repeatedly identified two dimensions of political ideology, rather than a single left-right spectrum. In social psychology, studies of human values have revealed two primary value dimensions underlying political ideology. @Rokeach1973 first classified these as _equality_ values (i.e., valuing brotherhood and equal opportunity) and _freedom_ values (i.e., valuing independence and free choice within groups). Subsequent work elaborated this two-dimensional value framework, referring to _communalism_ and _individualism_ values [@Katz1988], _self-transcendence_ and _conservation_ values [@Schwartz1992], and _international harmony_ and _national strength and order_ values [@Braithwaite1994]. Studies have also uncovered the two-dimensional structure of political attitudes from ratings of dictionary-based "isms", labelling these attitudinal dimensions as _unmitigated self-interest_ and _tradition-oriented religiousness_ [@Saucier2000]. In moral psychology, two primary clusters of moral concerns have been shown to predict political attitudes: _individualising moral foundations_, capturing ethics regarding care and harm, and _binding moral foundations_, capturing ethics regarding ingroup loyalty, deference to authority, and purity [@Graham2009]. In political psychology, researchers have found support for a two-factor model of political ideology which consists of the constructs _social dominance orientation_ (SDO), a preference for group-based hierarchy, and _right-wing authoritarianism_ (RWA), a preference for authoritarian aggression and submission [@Duckitt2009]. More recently, political psychologists have noted the regular clustering of policy views into _economic_ views (e.g., views on income redistribution, taxation, and welfare) and _social_ views (e.g., views on abortion and religious instruction in schools) [@Feldman2014; @Johnston2020]. Although these various pairs of constructs are psychologically distinct, they share many common elements, suggesting that research has independently converged upon two common underlying dimensions of political ideology.

Evolutionary theorists have attempted to explain the existence of the two dimensions of political ideology in various ways. For example, political egalitarianism has been shown to covary with upper-body strength and physical dominance [@Petersen2019], whereas authoritarianism and traditionalism have been linked to the behavioural immune system [@Tybur2016], negativity bias [@Hibbing2014], and adherence to social norms [@Kessler2008]. While these evolutionary approaches have illuminated aspects of each dimension separately, they do not attempt to explain the existence of a two-dimensional ideological structure from first principles.

## The dual evolutionary foundations of political ideology

Since politics is fundamentally the process by which humans deal with the myriad conflicts that arise from group living [@Haidt2012; @Petersen2015], any complete evolutionary account of the two-dimensional structure of political ideology must be consistent with research on the evolution of human group living. In fact, evidence from anthropology, primatology, and developmental psychology suggests that human group living evolved in two key steps [@Tomasello2012; @Tomasello2013; @Tomasello2014; @Jensen2014; @Sterelny2016]. In the first key step, humans became obligate cooperators, responding to resource scarcity by cooperating across wider interdependent networks and sharing the spoils of their cooperation equally. In the second key step, humans became group-minded, delineating their group’s boundaries with cultural markers and adhering to and enforcing group-wide social norms. Throughout the Pleistocene, these two key steps are thought to have paved the way from small kin bands to larger, more complex communities of hunter-gatherers.

These two primary responses to the challenges of human group living --- cooperation and group conformity --- can explain the existence of two foundational dimensions of political ideology that have been repeatedly uncovered in the social sciences [@Claessens2020]. The first dimension is broadly concerned with the trade-off between cooperative egalitarianism and competitive hierarchy. For example, equality and self-transcendence values reflect altruistic egalitarian motives, whereas unmitigated self-interest and SDO reflect competitive motives that promote inequality. The second dimension is broadly concerned with the trade-off between individualism and group conformity. For example, freedom and individualism values reflect independence from social control, whereas binding moral foundations and RWA reflect conformity to group-wide social norms. Together, these domains of cooperation and group conformity are predicted to shape human political psychology, not just in Western liberal democracies, but also in non-Western and small-scale societies.

This dual evolutionary theory, fully outlined in @Claessens2020, brings together and intersects with a diverse array of material, combining independent lines of evidence from multiple disciplines. But more work is required to develop and test the theory. In this paper, we provide the first test of one central prediction of the theory: that individual differences in general behavioural preferences for cooperation and group conformity can explain why people vary along the two dimensions of political ideology [@Claessens2020]^[We define cooperative behaviour as behaviour that provides a benefit to another individual at a personal cost [@West2007] and conformist behaviour as reliance on socially acquired information and adherence to and enforcement of social norms within cultural groups.]. Individual differences in behavioural preferences for cooperation and group conformity are thought to arise from a combination of biological predispositions and responses to environmental conditions. Based on this central prediction, an increased general behavioural preference for cooperation beyond close kin results in greater support for policies that promote large-scale impersonal cooperation and a more egalitarian distribution of resources, such as income redistribution, taxation, welfare, and pro-environmentalism. An increased general behavioural preference for group conformity results in greater support for policies that promote in-group conformity, norm enforcement, and the interests of the in-group, such as religious instruction in schools, capital punishment, and military spending.

## Previous research in behavioural economics

If general behavioural preferences for cooperation and group conformity do underlie differences in political opinion in modern humans, such differences should be manifest in preferences measured by domain-general incentivised behavioural tasks that involve decision-making with salient monetary rewards. Such tasks allow researchers to study variation in enduring social preferences without succumbing to issues of social desirability and experimenter demand that plague self-report methods [@Chaudhuri2009; @Gawronski2015; @Pisor2020].

@@

## The current study

Here, we extend prior work by combining a large suite of incentivised behavioural tasks with comprehensive survey data to test whether and to what extent cooperative and conformist behavioural preferences covary with measures of the two dimensions of political ideology. Across two pre-registered studies, we asked participants to complete a series of incentivised behavioural tasks measuring cooperative preferences, such as trust, altruism, and collective action, and conformist preferences, such as norm-enforcing punishment, social information use, and rule following. In Study 1, we used a battery of one-shot anonymous economic games, selected because they have previously been shown to capture two dimensions of behaviour, one reflecting a "cooperative phenotype" and another labelled "norm-enforcing punishment" [@Peysakhovich2014]. In Study 2, we extended this suite of games to include incentivised tasks capturing two further elements of conformist behaviour: social information use and rule following. In both studies, we used abstract, context-free behavioural tasks to best capture the general context-free behavioural preferences for cooperation and group conformity posited by our theory.

@@

Participants completed eight economic games, conducted online in real-time using oTree software (code available at https://osf.io/dwx8g/) [@Claessens2021; @Chen2016]. In an attempt to replicate previous work, these games are largely identical to those from @Peysakhovich2014.

@@

All games involved one-shot decisions between multiple anonymous players. Such one-shot anonymous games remove the well-documented effects of communication and known players on behaviour [@Chaudhuri2009; @Chaudhuri2011] and avoid potential confounds arising from repeated play, given that repeated game play may result in "super-games" and consequent cooperation even in finite repetitions [@Kreps1982]. The strategy method, in which participants provide responses for all possible roles in the games, was used to measure responses. This approach avoids feedback from other players and elicits similar behaviour to sequential play [@Brandts2000].

@@

Having confirmed the existence of a two-factor structure underlying gameplay, we next fitted pre-registered structural equation models to test the relationships between the two latent variables and political ideology. We simultaneously regressed the cooperation and punishment latent variables onto mean scores of six Likert-scale items for SDO (Cronbach's $\alpha$ = `r printnum(readd(alphaSDO))`) and RWA ($\alpha$ = `r printnum(readd(alphaRWA))`). In line with our second pre-registered hypothesis, we found that SDO significantly negatively predicted the cooperation latent variable with a small-to-medium effect size (standardised $\beta$ = `r printnum(standardizedSolution(sem1)[19,5])`, unstandardised _b_ = `r printnum(parameterEstimates(sem1)[19,5])`, 95% confidence interval [`r printnum(parameterEstimates(sem1)[19,9])`, `r printnum(parameterEstimates(sem1)[19,10])`], _p_ `r printp(parameterEstimates(sem1)[19,8])`, semi-partial _r_ = `r printnum(r_SDO_coop)`; Figure \@ref(fig:coopPunPlot)a). SDO also significantly positively predicted the punishment latent variable with a small effect size, which we did not hypothesise ($\beta$ = `r printnum(standardizedSolution(sem1)[21,5])`, _b_ = `r printnum(parameterEstimates(sem1)[21,5], digits=3)` [`r printnum(parameterEstimates(sem1)[21,9], digits=3)`, `r printnum(parameterEstimates(sem1)[21,10], digits=3)`], _p_ = `r printp(parameterEstimates(sem1)[21,8])`, semi-partial _r_ = `r printnum(r_SDO_pun)`), though the magnitude of this association was not as strong as with the cooperation latent variable (difference in absolute unstandardized coefficients = `r printnum(parameterEstimates(sem1)[59,5], digits=2)`, SE = `r printnum(parameterEstimates(sem1)[59,6], digits=3)`, p `r printp(parameterEstimates(sem1)[59,8])`). In contrast with our third pre-registered hypothesis, we found no significant relationship between RWA and the punishment latent variable ($\beta$ = `r printnum(standardizedSolution(sem1)[22,5])`, _b_ = `r printnum(parameterEstimates(sem1)[22,5], digits=3)` [`r printnum(parameterEstimates(sem1)[22,9], digits=3)`, `r printnum(parameterEstimates(sem1)[22,10], digits=3)`], _p_ = `r printp(parameterEstimates(sem1)[22,8])`, semi-partial _r_ = `r printnum(r_RWA_pun)`; Figure \@ref(fig:coopPunPlot)b). RWA was also unrelated to the cooperation latent variable ($\beta$ = `r printnum(standardizedSolution(sem1)[20,5])`, _b_ = `r printnum(parameterEstimates(sem1)[20,5], digits=2)` [`r printnum(parameterEstimates(sem1)[20,9], digits=2)`, `r printnum(parameterEstimates(sem1)[20,10], digits=2)`], _p_ = `r printp(parameterEstimates(sem1)[20,8])`, semi-partial _r_ = `r printnum(r_RWA_coop)`). This pattern of results held when additionally controlling for age, gender, ethnicity, education, socio-economic status, local deprivation, and religiosity (Supplementary Table \@ref(tab:semTable)).

@@

To further investigate the relationships between political ideology and gameplay, we conducted a series of exploratory linear regressions with mean SDO and RWA scores as outcome variables and individual game decisions as predictors, accounting for multiple comparisons and statistically controlling for demographics and game comprehension. We found that most cooperative decisions were negatively associated with SDO (Supplementary Figure \@ref(fig:indGamesPlotSDO)). Holding all other variables constant, individuals with higher SDO scores gave less in the Dictator Game, returned less in the Trust Game, contributed less in the Public Goods Game, did not coordinate in the Stag Hunt Game, offered less in the Ultimatum Game, and were more likely to take in the Third-Party Punishment Game. The effect sizes for these relationships were small, but consistent across games (semi-partial _r_ = `r min(rs)`-`r max(rs)`). Contrary to our predictions, SDO was also positively related to both "social" and "antisocial" punishment decisions in the Second-Party Punishment Game and Stag Hunt Game with Punishment (i.e., costly punishment of defectors and cooperators, respectively). Fewer individual game decisions predicted RWA (Supplementary Figure \@ref(fig:indGamesPlotRWA)). Holding all other variables constant, individuals with higher RWA scores were more likely to take in the Third-Party Punishment Game and engaged in greater anti-social punishment in the Second-Party Punishment Game and the Stag Hunt Game with Punishment.

@@

We did not find the predicted association between the punishment latent variable and RWA, nor did we find any link between the punishment latent variable and socially conservative policy views. One explanation for this lack of relationship is that while authoritarians might desire the sanctioning of norm violators, they would rather this punishment come from a third-party institution that they perceive to be legitimate than from an interpersonal encounter. Indeed, RWA items refer to sanctioning from strong leaders or governments, but rarely from individuals themselves [@Duckitt2009]. Another explanation is that the punishment games used in this study do not reflect general group conformist preferences as expected. As has been suggested previously [@Price2002; @Raihani2019], punishment in economic games may more closely track a competitive motive to increase one's relative payoffs over others, especially when the economic games are one-shot games with anonymous others in which there is no possibility for signalling, reputation building, or future behaviour modification.

@@

The materials for Study 2 were mostly identical to Study 1. As in Study 1, eight incentivised behavioural tasks were conducted online in real-time using oTree software (code available at https://osf.io/dwx8g/) [@Claessens2021; @Chen2016]. The strategy method was used to elicit responses in all possible roles, and participants played for points which were converted to New Zealand dollars at the same rate as in Study 1 (1 point = $0.035). To collect data continuously alongside the longitudinal New Zealand Attitudes and Values Study, the six tasks from previous work [@Peysakhovich2014] --- the Dictator Game, Trust Game, Public Goods Game, Ultimatum Game, Third-Party Punishment Game, and Second-Party Punishment Game --- were measured again (see Supplementary Table \@ref(tab:testRetestTable) for test-retest reliability). However, in Study 2, the coordination games were replaced with two different tasks.

The first task was the Berlin Estimate AdjuStment Task (BEAST) [@Molleman2019], a task designed to measure individual differences in conformity in the form of social information use. In this task, over five rounds, participants are required to report the number of animals present in briefly displayed images (see Supplementary Figure \@ref(fig:beastImages)). There are two parts to each round. In the first part, the image is displayed for six seconds and participants give their initial estimate of the number of animals ($E_1$). In the second part, participants are then made aware of a previous participant's estimate for the same image ($s$) and then give a second estimate ($E_2$). Previous participants' estimates are determined by searching a database of real estimates from 100 previous participants sampled by @Molleman2019. The computer searches for an estimate in this database that is (1) in the direction of the true answer (e.g., if the participant guesses 50 and the true answer is 60, social information will be higher than 50) and (2) deviates from the participant's first estimate according to a parameter $\Delta$: $s = E_1 * (1 \pm \Delta)$. Following the original paper, this $\Delta$ parameter varied across the five rounds, taking values of 0.25, 0.15, 0.20, 0.15, and 0.25 respectively. Participants were paid for their accuracy by randomly choosing a round and determining the distance between their estimate and the correct answer: 100 points - (distance * 5).

@@

In sum, we have shown clear differences in social preferences across the two dimensions of political ideology. We found that a general cooperative preference covaried with SDO and economic policy views, and informational conformity covaried with RWA and social policy views. These psychological differences were largely consistent with predictions of the dual evolutionary framework of political ideology [@Claessens2020]. The predictions not supported by the data suggest that the relationship between informational conformity and RWA and social policy views may not extend to conformity in the form of norm-enforcement and rule following, although demonstrating this conclusively will require further research using alternative behavioural measures. Taken together, our findings reveal the very different social preferences underlying both ideological dimensions, highlighting the importance of work that moves beyond unidimensional scales to elucidate the dual evolutionary foundations of political ideology.

@@

This work was supported by a Royal Society of New Zealand Marsden grant (UOA1711). The New Zealand Attitudes and Values Study is funded by a grant from the Templeton Religion Trust (TRT-2021-10418).

@@
